How to improve PCA based methods of genome scan using ecological data: detecting selection using RDA.

### 2017-03-15 article_genome_scan.tex
update

where $\beta_{jk}$ is the regression coefficient corresponding to the j-th SNP regressed by the k-th ordination axis, and $\epsilon_{j}$ is the residuals vector. To summarize the result of the regression analysis for the j-th SNP, we return a vector of z-scores $z_{j} = (z_{j1}, ..., z_{jK})$ where $z_{jk}$ corresponds to the z-score obtained when regressing the j-th SNP by the k-th ordination axis.
The test statistic is a robust Mahalanobis distance D computed using covRob function of the robustR package. We retain $K=5$ axis to compute Mahalanobis distances as it seems to explain most of the variance. D should be Khi2 distributed after a correction with inflation factor (Luu et al., 2016). Pvalues are computed using K degree of freedom. We use the FDR approach to control for false positives. Qvalue are computed with qvalueR package and a loci is considered as an outlier if its qvalue is less than 10\%.
For the analysis of simulated dataset (see below), we retain the first four ordination axis to compute Mahalanobis distances  as they seem to explain most of the variance in the data. To peform the ordination, we use the 10th environmental variables as input in the explanatory matrix. In the following example, we don't use phenotypic informations since these informations are often laking in environmental genomics. Neither we use geographical coordinates $(i,j)$ which is sometimes added to control for the geographical covariation in the differentiation pattern \citep{Frichot2013}.

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.1 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 20 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.

>>>>

where $\beta_{jk}$ is the regression coefficient corresponding to the j-th SNP regressed by the k-th ordination axis, and $\epsilon_{j}$ is the residuals vector. To summarize the result of the regression analysis for the j-th SNP, we return a vector of z-scores $z_{j} = (z_{j1}, ..., z_{jK})$ where $z_{jk}$ corresponds to the z-score obtained when regressing the j-th SNP by the k-th ordination axis.
The test statistic is a robust Mahalanobis distance D computed using covRob function of the robustR package. We retain $K=5$ axis to compute Mahalanobis distances as it seems to explain most of the variance. D should be Khi2 distributed after a correction with inflation factor (Luu et al., 2016). Pvalues are computed using K degree of freedom. We use the FDR approach to control for false positives. Qvalue are computed with qvalueR package and a loci is considered as an outlier if its qvalue is less than 10\%.
For the analysis of simulated dataset (see below), we retain the first five ordination axis to compute Mahalanobis distances as they seem to explain most of the variance in the data. To peform the ordination, we use the 10th environmental variables as input in the explanatory matrix. In the following example, we don't use phenotypic informations since these informations are often laking in environmental genomics. Neither we use geographical coordinates which is sometimes added to control for the geographical covariation in the differentiation pattern \citep{Frichot2013}.

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.1 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{fig:simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 20 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.
